\vspace{-3em} \textbf{ Mapping stellar surfaces\\ I: Degeneracies in the rotational light curve problem }

### 2021-01-29 paper1/ms.tex
softened language about spot contrast

\begin{abstract}
    %
    Thanks to missions like \emph{Kepler} and \emph{TESS}, we now
    have access to tens of thousands of high precision, fast
    cadence, and long baseline stellar photometric observations.
    %
    In principle, these light curves encode a vast amount of information about
    stellar variability and, in particular, about the distribution of
    starspots and other features on their surfaces.
    %
    Unfortunately, the problem of inferring stellar surface properties
    from a rotational light curve is famously ill-posed,
    as it often does not admit a unique solution.
    %
    Inference about the number, size, contrast, and location of spots
    can therefore depend very strongly on the assumptions of the model,
    the regularization scheme, or the prior.
    %
    The goal of this paper is twofold:
    (1) to explore the various degeneracies affecting the stellar
    light curve ``inversion'' problem and their effect on
    what can and cannot be learned from a stellar surface
    given unresolved photometric measurements; and
    (2) to motivate ensemble analyses of the light curves of
    many stars at once as a powerful data-driven alternative to common
    priors adopted in the literature.
    %
    We further derive novel results on the dependence of the null space
    on stellar inclination and limb darkening and
    show that single-band photometric measurements cannot uniquely
    constrain quantities like the spot contrast or the total
    spot coverage without the use of strong priors.
    %
    This is the first in a series of papers devoted to the development
    of novel algorithms and tools for the analysis of
    stellar light curves and spectral time series, with the
    explicit goal of enabling statistically robust inference
    about their surface properties.
\end{abstract}

@@

To summarize, the mean and variance of the ensemble of stellar light curves
in fractional units is
%
\begin{align}
    \mu      & = 1 - c \, n \, \alpha
    \nonumber                                     \\
    \sigma^2 & = c^2 \, n \, (\beta^2 - \alpha^2)
    \quad,
\end{align}
%
for some complicated functions $\alpha$ and $\beta$ of the distribution of
stellar inclinations,
rotation periods, and starspot properties.
%
If our observations were collected in these fractional units,
we could uniquely infer the spot contrast $c$ and the number of spots $n$,
since these enter as $c \, n$ and $c^2 n$ in the expressions for the mean
and variance, respectively, and these are straightforward statistics to
compute from the ensemble.
%
However, because of observations are made in \emph{relative} units, in which
we typically normalize to the mean, the amplitudes of features in light curves
can only tell us about the \emph{ratio}
%
\begin{align}
    \label{eq:ratio}
    \frac{\sigma}{\mu}
     & \propto \frac{c \sqrt{n}}{1 - c n \alpha}
\end{align}
%
for some value of $\alpha$.
In other words, even photometric ensemble analyses cannot tell
us about the values of the contrast and the number of spots independently.

A direct consequence of this normalization degeneracy is that it is not
possible to uniquely constrain the total spot coverage of a star from
single-band photometry without strong prior assumptions.
The total spot coverage $f_S$
is simply the (average) area of a spot
times the total number of spots divided by the total area of the sphere, which
may be expressed as
%
\begin{proof}{test_fS}
    \label{eq:fS}
    f_S = \frac{1}{2}\left(1 - \left<\cos r\right>\right)n
    \quad,
\end{proof}
%
given an angular spot radius $r$.
While $r$ may be uniquely constrained from the covariance structure of the data
\citepalias{PaperII}, $n$ cannot.

The arguments above are heuristic and based on only the first two moments
of the distribution of light curves in an ensemble.
It is possible, at least in principle, that higher order moments of the data
could encode information to
break the $c-N$ degeneracy, but these are in general much more difficult
to constrain from the data. It is also possible that we could place a
\emph{lower limit} on the spot contrast based on the normalized light
curve. In Equation~(\ref{eq:fapprox}) we implicitly assumed that
spots are allowed to overlap; under this assumption, it is possible to
double the contrast of a spot by simply adding another spot on top
of it. We could therefore generate light curves with arbitrarily
large dips by choosing a small value for $c$ and a very large value for $n$.
In reality, spots do not behave in this way: many overlapping spots
likely still have the same effective temperature! Therefore,
assuming there are no bright spots,
the maximum depth of a feature in the light curve (even in relative units)
could be used to place a lower limit on the spot contrast.

@@

In practice, this degeneracy usually manifests as a nonlinear correlation
between the contrast $c$ of a spot and the total number $n$
of spots on the surface of the star. Even in ensemble analyses,
one cannot in general learn these two quantities independently from
the data alone,
only a (complicated) function of the two (Equation~\ref{eq:ratio}).
%
A direct consequence of this degeneracy is that \textbf{constraints on
    the total spot coverage of a star from single-band light curves
    depend strongly on the model assumptions.}

@@

A better approach may be to collect photometric data in
multiple wavelength bands, an idea that has
been explored recently \citep[e.g.,][]{Gully2017,Guo2018}. Assuming the
locations and sizes of a star's spots are constant in wavelength, the amplitude
of the light curve in different bands (and in particular its slope as a function
of wavelength) can be used to directly constrain the
temperature, and hence the contrast, of the spots. This effectively breaks the
$c-N$ degeneracy. In practice, the effective size of spots may be different
at different wavelengths, which could complicate this picture somewhat, but
the point still stands that light curves collected in multiple bands
contain at least \emph{partial} information about the correct normalization.
%
A more detailed analysis of the information content of multi-band light curves,
and, by extension, spectroscopic timeseries, is deferred to a future paper
in this series.

>>>>

\begin{abstract}
    %
    Thanks to missions like \emph{Kepler} and \emph{TESS}, we now
    have access to tens of thousands of high precision, fast
    cadence, and long baseline stellar photometric observations.
    %
    In principle, these light curves encode a vast amount of information about
    stellar variability and, in particular, about the distribution of
    starspots and other features on their surfaces.
    %
    Unfortunately, the problem of inferring stellar surface properties
    from a rotational light curve is famously ill-posed,
    as it often does not admit a unique solution.
    %
    Inference about the number, size, contrast, and location of spots
    can therefore depend very strongly on the assumptions of the model,
    the regularization scheme, or the prior.
    %
    The goal of this paper is twofold:
    (1) to explore the various degeneracies affecting the stellar
    light curve ``inversion'' problem and their effect on
    what can and cannot be learned from a stellar surface
    given unresolved photometric measurements; and
    (2) to motivate ensemble analyses of the light curves of
    many stars at once as a powerful data-driven alternative to common
    priors adopted in the literature.
    %
    We further derive novel results on the dependence of the null space
    on stellar inclination and limb darkening and
    show that single-band photometric measurements cannot uniquely
    constrain quantities like the total
    spot coverage without the use of strong priors.
    %
    This is the first in a series of papers devoted to the development
    of novel algorithms and tools for the analysis of
    stellar light curves and spectral time series, with the
    explicit goal of enabling statistically robust inference
    about their surface properties.
\end{abstract}

@@

To summarize, the mean and variance of the ensemble of stellar light curves
in fractional units is
%
\begin{align}
    \mu      & = 1 - c \, n \, \alpha
    \nonumber                                     \\
    \sigma^2 & = c^2 \, n \, (\beta^2 - \alpha^2)
    \quad,
\end{align}
%
for some complicated functions $\alpha$ and $\beta$ of the distribution of
stellar inclinations,
rotation periods, and starspot properties.
%
If our observations were collected in these fractional units,
we could uniquely infer the spot contrast $c$ and the number of spots $n$,
since these enter as $c \, n$ and $c^2 n$ in the expressions for the mean
and variance, respectively, and these are straightforward statistics to
compute from the ensemble.
%
However, because of observations are made in \emph{relative} units, in which
we typically normalize to the mean, the amplitudes of features in light curves
can only tell us about the \emph{ratio}
%
\begin{align}
    \label{eq:ratio}
    \frac{\sigma}{\mu}
     & \propto \frac{c \sqrt{n}}{1 - c n \alpha}
\end{align}
%
for some value of $\alpha$.
In other words, even photometric ensemble analyses may not be able to tell
us about the values of the contrast and the number of spots independently.

A direct consequence of this normalization degeneracy is that it may not be
possible to uniquely constrain the total spot coverage of a star from
single-band photometry without strong prior assumptions.
The total spot coverage $f_S$
is simply the (average) area of a spot
times the total number of spots divided by the total area of the sphere, which
may be expressed as
%
\begin{proof}{test_fS}
    \label{eq:fS}
    f_S = \frac{1}{2}\left(1 - \left<\cos r\right>\right)n
    \quad,
\end{proof}
%
given an angular spot radius $r$.
While $r$ may be uniquely constrained from the covariance structure of the data
\citepalias{PaperII}, $n$ cannot.

The arguments above are heuristic and based on only the first two moments
of the distribution of light curves in an ensemble.
It is possible, at least in principle, that higher order moments of the data
could encode information to
break the $c-n$ degeneracy, but these are in general more difficult
to constrain from the data. It is also possible that we could place a
\emph{lower limit} on the spot contrast based on the normalized light
curve. In Equation~(\ref{eq:fapprox}) we implicitly assumed that
spots are allowed to overlap; under this assumption, it is possible to
double the contrast of a spot by simply adding another spot on top
of it. We could therefore generate light curves with arbitrarily
large dips by choosing a small value for $c$ and a very large value for $n$.
In reality, spots do not behave in this way: many overlapping spots
likely still have the same effective temperature! Therefore,
assuming there are no bright spots,
the maximum depth of a feature in the light curve (even in relative units)
could be used to place a lower limit on the spot contrast.
Moreover, in \citetalias{PaperII} we empirically show that, depending on the choice
of prior, ensemble analyses may be able to place stronger constraints on
the contrast $c$ than on the number of spots $n$.

@@

In practice, this degeneracy usually manifests as a nonlinear correlation
between the contrast $c$ of a spot and the total number $n$
of spots on the surface of the star. Even in ensemble analyses,
one cannot in general learn these two quantities independently from
the data alone,
only a (complicated) function of the two (Equation~\ref{eq:ratio}).
In \citetalias{PaperII} we show empirically that
careful analysis of the covariance structure of ensembles of light curves
may shed some light on $c$ but cannot uniquely constrain $n$.
%
A direct consequence of this degeneracy is that \textbf{constraints on
    the total number of spots or on the spot coverage of a star from single-band light curves
    depend strongly on the model assumptions.}

@@

A better approach may be to collect photometric data in
multiple wavelength bands, an idea that has
been explored recently \citep[e.g.,][]{Gully2017,Guo2018}. Assuming the
locations and sizes of a star's spots are constant in wavelength, the amplitude
of the light curve in different bands (and in particular its slope as a function
of wavelength) can be used to directly constrain the
temperature, and hence the contrast, of the spots. This effectively breaks the
$c-n$ degeneracy. In practice, the effective size of spots may be different
at different wavelengths, which could complicate this picture somewhat, but
the point still stands that light curves collected in multiple bands
contain at least \emph{partial} information about the correct normalization.
%
A more detailed analysis of the information content of multi-band light curves,
and, by extension, spectroscopic timeseries, is deferred to a future paper
in this series.
